---
title: "Methods used in `dsrci` to construct confidence intervals for Directly Standardized Rates."
author: "Michael Nelson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  \VignetteIndexEntry{Methods used in `dsrci` to construct confidence intervals for Directly Standardized Rates.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Notation

Let $x_1, \ldots, x_k$ be the cell counts  with associated random variables $X_1, \ldots, X_n$, $n_1, \ldots, n_k$ be the number of person-years for each cell and assume $X_i~\mathcal{P}\left(\theta_i\right)$.

Let $c_i, \ldots, c_k$ be the number of person-years in a standard population. 

Let the weight for the $i$th cell be

$$
w_i = \frac{c_i}{n_i\left(\sum_{j=1}^k{c_j}\right)}
$$
The directly standardized rate (DSR) can be calculated as

$$
y = \sum_{i=1}^k w_ix_i
$$
An estimate of the variance of the DSR is

$$
\upsilon = \sum_{i=1}^k w_i^2 x_i
$$
Let $Y = \sum_{i=1}^k w_i X_i$, and $E\left[Y\right] = \mu = \sum_{i=1}^k w_i \theta_i$. 


# Constructing confidence intervals

Fay & Feuer (1997), Tiwari et al (2006) and Ng et al (2008) describe and compare a number of confidence interval methods.

Throughout, let $L\left(y\right), U\left(y\right)$ be the lower  and upper bounds of a $100(1-\alpha)\%$ confidence interval.

The example data from `example(wspoissonTest, package = 'asht')`

```{r init}
library(dsrci)
## from example(wspoissonTest, package = 'asht')
## birth data on Down's syndrome from Michigan, 1950-1964
## see Table II  of Fay and Feuer (1997)
## xfive= counts for mothers who have had 5 or more children
## nfive and ntotal are number of live births 
xfive <- c(0, 8, 63, 112, 262, 295)
nfive <- c(327, 30666, 123419, 149919, 104088, 34392)
ntotal <- c(319933, 931318, 786511, 488235, 237863, 61313)
```

## Asymptotic method `ci.method = "asymptotic"`

### Normal approxmation (`trans = "none"`)

```{r asymptotic-none}
# ci.method = "asymptotic", trans = "none"
# are the default values
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95)
# you can also call ci.asymptotic directly, 
# only when weights = weights not standard population counts
# this returns the results without applying a multiplicative
# factor
ci.asymptotic(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95)
```


### Log transformation (`trans = "log"`)

```{r asymptotic-log}
# ci.method = "asymptotic", trans = "log"
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, trans = "log")
# using ci.asymptotic
ci.asymptotic(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, trans = "log")
```

### Cube root (`trans = "cube.root"`)
Using the delta method, 


```{r asymptotic-cube}
# ci.method = "asymptotic", trans = "cube.root"
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, trans = "cube.root")
# using ci.asymptotic
ci.asymptotic(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, trans = "cube.root")
```

### Edgeworth correction for skew (`trans = "skew"`)
```{r asymptotic-skew}
# ci.method = "asymptotic", trans = "skew"
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, trans = "skew")
# using ci.asymptotic
ci.asymptotic(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, trans = "skew")
```


## Moment matching (`ci.method = "moments"`)

Following Dobson et al (1991), an approximate confidence interval can be obtained as a linear function of the confidence interval for a single Poisson paramenter ($X = \sum_{i=1}^k X_i$) where the confidence interval for this *unweighted* sum of poisson parameters $\sum_{i=1}^k \theta_i$ is $\left(X_L,X_U \right)$.

An approximate confidence interval for the *weighted* sum of $\theta$ is

$$
T_L = Y + \sqrt{\frac{\upsilon}{y}}\left(X_L - X\right)
$$
$$
T_U = Y + \sqrt{\frac{\upsilon}{y}}\left(X_U - X\right)
$$

A number of methods for estimating $\left(X_L,X_L\right)$ are implemented in `dsrci`

### Dobson (exact) (`type = "dobson"`)
```{r moments-dobson}
# ci.method = "moments", type = "dobson" 
# note type = "dobson" is the default
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "moments", type = "dobson")
# using ci.moments
ci.moments(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95)
```

### Boise-Monson (`type = "boise.monson"`)


```{r moments-boise}
# ci.method = "moments", type = "boise.monson" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "moments", type = "boise.monson")
# using ci.moments
ci.moments(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "boise.monson")
```


### Normal approximation (`type = "normal"`)

```{r moments-normal}
# ci.method = "moments", type = "normal" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "moments", type = "normal")
# using ci.moments
ci.moments(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "normal")
```

### Wilson-Hilferty (`type = "wilson.hilferty"`)

```{r moments-wilson}
# ci.method = "moments", type = "wilson.hilferty" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "moments", type = "wilson.hilferty")
# using ci.moments
ci.moments(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "wilson.hilferty")
```

### Byar's method (`type = "byar"`)


```{r moments-byar}
# ci.method = "moments", type = "byar" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "moments", type = "byar")
# using ci.moments
ci.moments(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "byar")
```


### Exact mid-*p* confidence interval

```{r moments-midp}
# ci.method = "moments", type = "midp" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "moments", type = "midp")
# using ci.moments
ci.moments(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "midp")
```


### Approximation to mid-*p* confidence interval

```{r moments-approx.midp}
# ci.method = "moments", type = "approx.midp" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "moments", type = "approx.midp")
# using ci.moments
ci.moments(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "approx.midp")
```

### Simple Approximation to mid-*p* confidence interval


```{r moments-simple.midp}
# ci.method = "moments", type = "approx.midp" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "moments", type = "simple.midp")
# using ci.moments
ci.moments(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "simple.midp")
```

## Gamma distribution

All Gamma-distribution methods call `asht::wspoissonTest`

### Fay and Feuer (`type = "max")

```{r gamma-max}
# ci.method = "gamma", type = "max" (this is the default type)
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "gamma", type = "max")
# using ci.gamma
ci.gamma(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "max")
```

### Mid-*p* (`type = "midp"`)

```{r gamma-midp}
# ci.method = "gamma", type = "midp"
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "gamma", type = "midp")
# using ci.gamma
ci.gamma(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "midp")
```

### Tiwari et al 2006 `mean` (`type = "mean"`)

```{r gamma-mean}
# ci.method = "gamma", type = "mean" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "gamma", type = "mean")
# using ci.gamma
ci.gamma(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "mean")
```

### Tiwari et al 2006 `minmaxavg` (`type = "minmaxavg"`)

```{r gamma-minmaxavg}
# ci.method = "gamma", type = "minmaxavg" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "gamma", type = "minmaxavg")
# using ci.gamma
ci.gamma(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "minmaxavg")
```

### Tiwari et al 2006  (`type = "tcz"`)

```{r gamma-tcz}
# ci.method = "gamma", type = "tcz" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "gamma", type = "tcz")
# using ci.gamma
ci.gamma(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "tcz")
```

## Beta Distribution

### Tiwari et al 2006  (`type = "tcz"`)

```{r beta-tcz}
# ci.method = "beta", type = "tcz" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "beta", type = "tcz")
# using ci.beta
ci.beta(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "tcz")
```

### Tiwari et al 2006 Continuity Correction (`type = "cc"`)

```{r beta-cc}
# ci.method = "beta", type = "cc" 
dsr(x = xfive, n =  nfive, w = ntotal, mult = 1e5, level = 0.95, ci.method = "beta", type = "cc")
# using ci.beta
ci.beta(xfive, ntotal/(nfive*sum(ntotal)), level = 0.95, type = "cc")
```

# Comparison of Methods

```{r all-methods}
# a function to loop through methods and return a data.frame
loop_methods <- function(list, level = 0.95, x = xfive, n = nfive, w= ntotal, mult = 1e5, ci.method){
  # loop through "types"
  results <- lapply(list, dsr, level = level, x = x, w = w, n = n, mult = mult, ci.method = ci.method)
  # combine into a single data.frame
  results <- do.call(rbind, lapply(results, as.data.frame))
}
# transformations implemented for "asymptotic"
trans <- c("none", "log", "cube.root", "skew")
asymp.results <- loop_methods(trans, ci.method = "asymptotic")
# types implemented for "moments"
m.types <- c("dobson", "boise.monson", "normal", "wilson.hilferty", "byar", 
    "midp", "approx.midp", "simple.midp")
moments.results <- loop_methods(m.types, ci.method = "moments")
# types implemented for "gamma"
g.types <- c("max", "midp", "mean", "minmaxavg", "tcz")
gamma.results <- loop_methods(g.types, ci.method = "gamma")
# types implemented for "beta"
b.types <- c("tcz", "cc")
beta.results <- loop_methods(b.types, ci.method = "beta")
# there are no "types" for the approximate bootstrap
# method - no need to loop
boot.results <- as.data.frame(dsr(xfive, nfive, ntotal, mult = 1e5, ci.method = 'bootstrap'))
# combine into a single data.frame
all.results <- rbind(asymp.results, moments.results,
                     gamma.results, beta.results,
                     boot.results)
```

```{r show-results, results='asis'}
# show results
knitr::kable(all.results[,c("ci.method", "method.arg", "estimate", "lower", "upper")])
```

# References


Dobson, AJ, Kuulasmaa, K, Eberle, E and Scherer, J (1991)
'Confidence intervals for weighted sums of Poisson parameters',*Statistics in Medicine*, **10**: 457—462.

Swift, MB (1995). 'Simple confidence intervals for standardized rates based on the approximate bootstrap method', *Statistics in Medicine*, **14**, 1875—1888.

Fay & Feuer (1997). 'Confidence intervals for directly standardized rates: a method based on the gamma distribution.' *Statistics in Medicine*. **16**: 791-801.

Tiwari, Clegg, & Zou (2006). 'Efficient interval estimation for age-adjusted cancer rates.' *Statistical Methods in Medical Research* **15**: 547-569.

Ng, Filardo, & Zheng (2008). 'Confidence interval estimating procedures for standardized incidence rates.' *Computational Statistics and Data Analysis* **52**: 3501-3516.

Fay MP (2010). 'Two-sided Exact Tests and Matching Confidence Intervals for Discrete Data'.  *R Journal* **2**(1):53-58.

Fay & Kim (2017). 'Confidence intervals for directly standardized rates using mid-p gamma intervals.' *Biometrical Journal*. doi:10.1002/bimj.201600111.

Fay  MP (2017). '`asht`: Applied Statistical Hypothesis Tests'. *`R` package version 0.9.* https://CRAN.R-project.org/package=asht